Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-10-19 docs/manuscript/paper.tex
snapshot of more comments

Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$. After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames. Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{equation}
	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
	\label{eq:g_func}
\end{equation}
and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$. It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.

@@

\section{Validation procedure with synthetic data} \label{sec:testing}
In this section we outline how synthetic data is generated in order to validate the analysis scheme in Section \ref{sec:detect}. In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$. In Section \ref{sec:gendata} we demonstrate how to solve the equations of Section \ref{sec:model} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(n)}(t)$.

\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
\begin{figure}
	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work. The pulsar distances relative to the observer are also indicated, with a distance $\leq 2 \, {\rm kpc}$ for $38$ pulsars. The grey dashed curve denotes the Galactic plane.}
	\label{fig:pulsar_distrib}
\end{figure}
We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}. The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA. For each pulsar we adopt the fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter. A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility. The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}. The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist. The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency} and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$. The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
\begin{align}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
\end{align}
where $T_{\rm cad}$ is the cadence of the timing observations and for MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$, for $\pm 2\sigma$ confidence limits. The uncertainties on these parameters are broad; for the purposes of generating astrophysically representative synthetic data we straightforwardly take the best fit parameter value. Throughout this work we assume that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$. In order to relate Equation \eqref{eq:sigmap}, to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
\end{eqnarray}
For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline

As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
\begin{eqnarray}
	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
\end{eqnarray}
and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach. We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$. In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$. In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}. We see that the synthetic and empirical residuals are qualitatively similar. For this pulsar the synthetic residuals were generated using $\sigma^{(n)} = 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the value inferred from the noise model of Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, using the best fit parameters, $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$. This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars. If we set $\ln \alpha_1 = -38$ in Equation \eqref{eq:sigmap}, close to its lower limit, then the estimated $\sigma^{(n)}$ for  PSR J1939+2134 is $\sigma^{(n)} =3.90 \times 10^{-27}$ which is in much better agreement with the value inferred from the residuals comparison method. This discrepancy between the two estimates of $\sigma^{(n)}$ is particular to PSR J1939+2134; for all other pulsars we find the two estimates of $\sigma^{(n)}$ are comparable to within a few orders of magnitude.
\begin{figure}
	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}
	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J1939+2134. The actual residuals are sourced from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)} = 2\times 10^{-27}$ s$^{-3/2}$. The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}. We set the TOA uncertainty to be a constant $0.1 \mu$s. The blue and orange residuals are qualitatively similar by inspection. Similar results are obtained for the other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} by tuning the value of $\sigma^{(n)}$.}
	\label{fig:qualitative_compare}
\end{figure}
\subsection{Generating a synthetic sequence of pulse frequencies}\label{sec:gendata}
We generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
\begin{enumerate}[leftmargin=2em]
	\item Integrate Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}, to obtain random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}
\end{enumerate}
Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \eqref{sec:synt_pta}. For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence. The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, as
	\sigma_{\rm m} \approx f_{\rm p}^{(n)} \sigma_{\rm TOA} \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct; Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise. For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz. The most accurately timed pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$ Hz. Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred. Note that whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}$ is modified by a different realisation of the measurement noise for each pulsar. \newline

Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } \sigma^{(n)} \Delta W_t \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta W_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}). To fix the problem, we subtract the deterministic frequency evolution and track the new variable, equivalent to a change of variables:
\begin{equation}
	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ .
\end{equation}  
We similarly modify the measurement variable to be
\begin{equation}
	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ ,
\end{equation}
where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris which is measured to high accuracy in practice by {\sc tempo2}. For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately. Equation \eqref{eq:measurement} is then updated to read 
\begin{equation}
	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
	\label{eq:measurement_cov}
\end{equation}
We emphasise that this change of variables is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g. long double, quadruple). It does not remove any degrees of freedom nor does it involve an approximation. In particular $f_{\rm em}^{(n)}(t_1)$
and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, Equations \ref{eq:spinevol}--\eqref{eq:xieqn}.

\section{Representative example: single SMBHB source}\label{sec:rep_example}
In this Section we apply the analysis scheme of Section \ref{sec:detect} and the validation procedure of Section \ref{sec:testing} to a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source. We remind the reader that the analysis in this paper targets a single SMBHB source. The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper. In Section \ref{sec:priors} we outline the priors chosen on $\boldsymbol{\theta}$. In Section \ref{sec:parameter_estim} we apply the workflow of Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise. We go on in Section \ref{sec:multiple_noise} to extend this parameter estimation procedure to multiple realisations of the noise processes. In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$. \newline

The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are selected to be astrophysically reasonable and representative. The injected $\boldsymbol{\theta}_{\rm gw}$ are summarised in the ``Injected Values" column of Table \ref{tab:parameters_and_priors}. The static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$ are as described in Section \ref{sec:synt_pta} and are also shown in Table \ref{tab:parameters_and_priors}.

\subsection{Prior distributions}\label{sec:priors}
For the Bayesian methods used in this work, one must select reasonable priors, $\pi(\boldsymbol{\theta})$, for the complete set complete of static parameters. For $\pi(\boldsymbol{\theta}_{\rm gw})$ we choose standard non-informative priors \citep[e.g.][]{Bhagwat2021} as summarised in Table \ref{tab:parameters_and_priors}. The choice of $\pi(\boldsymbol{\theta}_{\rm psr})$ requires some additional discussion. \newline

>>>>

Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$. After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames. Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{equation}
	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
	\label{eq:g_func}
\end{equation}
and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$. It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.

@@

\section{Validation with synthetic data} \label{sec:testing}
In this section we outline how synthetic data are generated in order to validate the analysis scheme in Section \ref{sec:detect}. Synthetic data enable validation to occur systematically and under controlled conditions. In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$. In Section \ref{sec:gendata} we demonstrate how to solve Equations \eqref{eq:spinevol} -- \eqref{eq:frequency_evolution}, and \eqref{eq:measurement} -- \eqref{eq:g_func} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(n)}(t)$.

\subsection{Constructing a synthetic PTA}\label{sec:synt_pta}
\begin{figure}
	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
	\caption{Spatial distribution in Galactic coordinates of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work. The pulsar distances relative to the observer are also indicated, with a distance $\leq 2 \, {\rm kpc}$ for $38$ pulsars. The grey dashed curve denotes the Galactic plane.}
	\label{fig:pulsar_distrib}
\end{figure}
We consider, by way of illustration, a synthetic PTA composed of the 47 pulsars in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}. The NANOGrav pulsars are chosen arbitrarily as being representative of a typical PTA; the analysis below extends unchanged to any other PTA. \newline

For each pulsar we adopt fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter. A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility. The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}. The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. \newline

The remaining static pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$, for which no direct measurements exist. The ratio $\sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ sets the typical root mean square fluctuations in $f_p^{(n)}(t)$, as discussed in Section \ref{sec:psr_frequency}, and the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. We follow two complementary approaches in order to  set physically reasonable values for $\sigma^{(n)}$. The first approach relies on the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} which gives the standard deviation of the pulsar TOAs, $\sigma_{\rm TOA}^{(n)}$, empirically as
\begin{align}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)=& \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right) \nonumber \\ 
	&+ \alpha_4 \ln \left( \frac{T_{\rm cad}}{ 1 \text{ year}} \right) \ , \label{eq:sigmap}
\end{align}
where $T_{\rm cad}$ is the cadence of the timing observations. For MSPs the best fit parameters are $\ln \alpha_1 = -20 \pm 20$, $\alpha_2 = 1 \pm 2 $, $\alpha_3 = 2 \pm 1$, $\alpha_4 = 2.4 \pm 0.6$ ($\pm 2\sigma$ confidence limits). The uncertainties are broad; for the purpose of generating astrophysically representative synthetic data we straightforwardly adopt the central values in this paper. Throughout this work we assume for simplicity that all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$. In order to relate Equation \eqref{eq:sigmap} to $\sigma^{(n)}$ in Equation \eqref{eq:spinevol}, we equate $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week, obtaining
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
\end{eqnarray}
For the synthetic NANOGrav PTA depicted in Figure \ref{fig:pulsar_distrib}, the median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, with $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134 \newline

As a cross-check, we estimate $\sigma^{(n)}$ by solving Equation \eqref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}}, generating a synthetic phase solution,
\begin{eqnarray}
	\varphi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') \ ,
\end{eqnarray}
and adjusting $\sigma^{(n)}$ iteratively to generate phase residuals which qualitatively (i.e.\ visually) resemble empirical phase residuals measured from real pulsars; see Section 2 in \cite{Vargas} for a successful example of this approach. We obtain empirical phase residuals from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. A visual cross-check is sufficient for the purposes of this paper, where we seek broadly representative values for $\sigma^{(n)}$. In an analysis involving real PTA data, $\sigma^{(n)}$ would be estimated from the data jointly with the other static parameters in ${\boldsymbol{\theta}}$. In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for PSR J1939+2134, one of the 47 NANOGrav pulsars plotted in Figure \ref{fig:pulsar_distrib}. We see that the synthetic and empirical residuals are qualitatively similar. For this pulsar the synthetic residuals are generated using $\sigma^{(n)} = 2 \times 10^{-27}$ s$^{-3/2}$ which is $\ll$ than the central value $\sigma^{(n)} = 2.56 \times 10^{-19}$ s$^{-3/2}$ inferred from Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f}, but is consistent with the wide ranges on $\alpha_1$, $\dots$, $\alpha_4$, e.g.\ $\ln \alpha_1 = -38$ implies $\sigma^{(n)}= 3.90 \times 10^{-27}$ for PSR J1939$+$2134.

%This highlights how given the large uncertainties on the parameters of Equation \eqref{eq:sigmap} using the best fit values may be less appropriate for certain pulsars. If we set $\ln \alpha_1 = -38$ in Equation \eqref{eq:sigmap}, close to its lower limit, then the estimated $\sigma^{(n)}$ for  PSR J1939+2134 is $\sigma^{(n)} =3.90 \times 10^{-27}$ which is in much better agreement with the value inferred from the residuals comparison method.

\begin{figure}
	\includegraphics[width=\columnwidth]{images/example_residuals_plot2}
	\caption{Actual (left panel, blue points) and synthetic (right panel, orange points) phase residuals for NANOGrav millisecond pulsar PSR J1939+2134. The actual residuals are sourced from the NANOGrav 12.5 year wideband timing dataset \citep{pennucci_timothy_t_2020_4312887,nanogravwideband}. The synthetic residuals are generated by numerically solving Equation \eqref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)} = 2\times 10^{-27}$ s$^{-3/2}$. The error bars indicate the uncertainty in the phase residuals and are generated by propagating the uncertainty in the TOAs through {\sc tempo2}. We set the TOA uncertainty to be a constant, $0.1 \mu$s. The blue and orange residuals are qualitatively similar by inspection. Similar results are obtained for the other 46 pulsars in the synthetic PTA in Section \ref{sec:synt_pta} by tuning the value of $\sigma^{(n)}$.}
	\label{fig:qualitative_compare}
\end{figure}
\subsection{Generating a synthetic sequence of pulse frequencies}\label{sec:gendata}
We generate $N$ synthetic noisy time series of the measured pulse frequency $f_{\rm m}^{(n)}(t)$, one for each pulsar $1\leq n \leq N$, as follows:
\begin{enumerate}[leftmargin=2em]
	\item Integrate Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} numerically for the synthetic PTA described in Section \ref{sec:synt_pta}, to obtain random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$.
	\item Map from $f_{\rm p}^{(n)}(t)$ to $f_{\rm m}^{(n)}(t)$ via Equations \eqref{eq:measurement} and \eqref{eq:g_func}.
\end{enumerate}
Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified for the synthetic PTA outlined in Section \ref{sec:synt_pta}. For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence. \newline

The measurement noise covariance as defined in Equation \eqref{eq:vareps} can be approximately related to the uncertainty in the pulse TOA, $\sigma_{\rm TOA}$, through
\begin{equation}
	\sigma_{\rm m} \approx \sigma_{\rm TOA} f_{\rm p}^{(n)}  \ {T_{\rm cad}}^{-1} \ . \label{eq:sigma_m_eqn}
\end{equation}
Although Equations \eqref{eq:sigmap_f} and \eqref{eq:sigma_m_eqn} superficially resemble one another, they are distinct. Equation \eqref{eq:sigmap_f} deals with the timing noise intrinsic to the pulsar due to rotational irregularities, whereas Equation \eqref{eq:sigma_m_eqn} handles the detector measurement noise. For a millisecond pulsar with $f_{\rm p}^{(n)} \sim 100$ Hz observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$s, Equation \eqref{eq:sigma_m_eqn} implies $\sigma_{\rm m} \sim 10^{-10}$ Hz. The most accurately timed pulsars have $\sigma_{\rm TOA} \sim 10 $ ns and $\sigma_{\rm m} \sim 10^{-12}$ Hz. Throughout this paper we fix $\sigma_{\rm m} = 10^{-11}$ Hz and take it as known \textit{a priori} rather than a parameter to be inferred for the sake of simplicity; this assumption can be relaxed easily when analysing real data. Whilst $\sigma_{\rm m}$ is the same for every pulsar, $f_{\rm m} ^{(n)}(t)$ is constructed from a different random realisation of $\varepsilon^{(n)}(t)$ for each pulsar. \newline

Finite precision arithmetic leads to numerical errors when solving Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} $ \text{in  the regime } |\sigma^{(n)} \Delta B_t | \ll f_{\rm p}^{(n)}$ relevant to PTAs, where $\Delta B_t$ labels an increment of Brownian motion (see Appendix \ref{sec:kalman}). To fix the problem, we subtract the deterministic frequency evolution and track the new variable
\begin{equation}
	f_{\rm p}^{* (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)} \ , \label{eq:hetero1}
\end{equation}  
equivalent to a change of variables. We similarly modify the measurement variable to be
\begin{equation}
	f_{\rm m}^{* (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{\diamond (n)} \ , \label{eq:hetero2}
\end{equation}
where $ f_{\rm em}^{\diamond (n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris returned by {\sc tempo2}  which is measured to high accuracy in practice. For synthetic data we can set $ f_{\rm em}^{\diamond (n)} = f_{\rm em}^{(n)}$ without loss of generality, but this is impossible generally for astronomical observations, because the spin-down ephemeris is only known approximately. Equation \eqref{eq:measurement} is then updated to read 
\begin{equation}
	f_{\rm m}^{* (n)}(t) = f_{\rm p}^{* (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}(t-d)\left[ 1-g^{(n)}(t)\right] \ .
	\label{eq:measurement_cov}
\end{equation}
We emphasise that the change of variables in Equations \eqref{eq:hetero1} and \eqref{eq:hetero2} is a convenient device to bring the numerical values into a reasonable dynamic range without having to use excessively long floating point formats (e.g. long double, quadruple). It does not remove any degrees of freedom nor does it involve an approximation. In particular $f_{\rm em}^{(n)}(t_1)$
and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain as static parameters but appear in the measurement equation \eqref{eq:measurement_cov} rather than the dynamical state equations, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}.

\section{Representative end-to-end analysis} \label{sec:rep_example}
In this section we apply the analysis scheme in Section \ref{sec:detect} and the validation procedure in Section \ref{sec:testing} to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source. The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper. The goal of this section is to run end-to-end through every step in the analysis for a representative worked example to help the reader implement the scheme and reproduce the results in Sections \ref{sec:parameter_space}--\ref{sec:discussion}. To assist with reproducibility, we present intermediate outputs from each step. In Section \ref{sec:priors} we define the priors chosen on $\boldsymbol{\theta}$.

In Section \ref{sec:earth_psr_terms} we do this...

\textcolor{red}{TK to do}

In Section \ref{sec:parameter_estim} we apply the workflow in Section \ref{sec:methodsummary} to estimate $\boldsymbol{\theta}$, for a single realisation of the pulsar process noise and the measurement noise. In Section \ref{sec:multiple_noise} we extend the parameter estimation exercise to multiple noise realizations to quantify the irreducible ``cosmic'' variance in the parameter estimates. In Section \ref{sec:detection} we calculate the detectability of the source as a function of $h_0$. \newline
